### 2022-09-12 AISTATS2023/paper.tex
updated background

\paragraph{Notation.}

We refer to random variables (rv) using uppercase \( X \), whereas to
simple variables with plain lowercase \( x \). Bold denotes a vector;
\( \xb \) for simple variables or \(\mathbf{X}\) for rvs. Often, we
partition the input vector \(\xb \in \mathbb{R}^D\) to the feature of
interest \(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} PDPlots formulate the feature
effect of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
integration of local effects. The local effects are measured as the
expected change on the output
\( \frac{\partial f(x_s, \mathbf{X}_c)}{\partial x_s} \) over the
conditional distribution \( \Xcb|X_s\). The formula that
defines ALE is presented below:

\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = c + \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
The constant \(c\) is used for centering the ALE plot. PDPlots
integrate over unrealistic instances due to the use of the marginal
distribution \( p(\mathcal{X}_1) \). Therefore, they incorrectly
result in a quadratic effect in the region \(x_1 \in [0, 1]\). MPlots
resolve this issue using the conditional distribution
\( \mathcal{X}_2|\mathcal{X}_1 \) but suffer from computing combined
effects. ALE plots resolve, both they have two drawbacks TODO add more
info.

TODO: Discuss the interpretation of ALE. The differences from PDP. Analyze both methods as processes.

\paragraph{Qunatify the Heterogeneous Effects.} It is crucial for
feature effect methods to inform about the heterogeneous
effects. Elaborate. Interpretation of the heterogeneous effects behind
the global effect is available only for PDP, with three different
approaches; (a) ICE and d-ICE plots provide a visual understanding of
the heterogeneous effects. (b) grouping of ICE in homogeneous
clusters, for spliting the input space into subspace(s) with
homogeneous effects (c) Feature Interaction strength indexes, like
H-statistic, provide a value indicating how much a feature interacts
with the others (not the type of interaction). There is no method for
quantifying the heterogeneous effects, based on ALE. Therefore, no
method to exploit the advantages of ALE while, on the same time,
informing about the heterogeneous effects.

\paragraph{Bin Spliting.} ALE also has the peculiarity of spliting the
axis into intervals, allocating the instances of the training set in
the intervals and compute a single (constant) effect in each
interval. With DALE extension, bin spliting is decoupled from instant
effect estimation. With our extension for measuring the heterogeneous
effects, we transfrom interval spliting from a step to a clustering
problem with a meaningful objective to minimise. We provide a thorough
analysis, where we show that our objective has a consistent
meaning. It can be split in two parts; the first part is the
unavoidable uncertainty due to the natural characteristics of the
experiment, i.e.,~the data generating distribution and the black-box
function. The second part is an added uncertainty due to the
limited-samples estimation, that enforces to create groups with
constant main effect. We opt for minimizing the objective, i.e. sum of
the two uncertainties, that given that the first uncertainty is
independent of the bin spliting, therefore we want to minimize the
added uncertainty. To conclude, we transform the axis-spliting into an
unsupervised clustering problem with a principled objective. We a
computationally-grounded solution that works out-of-the-box, relaxing
the user from providing a hyperparameter without any indication which
one is the correct. This step can be used independently of whether the
user wants to explore the heterogeneous effects or not.

@@

ALE defines the local effect of \(x_s\) on \(f(\cdot)\) at a single
point\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are weigthed by the conditional distribution
\(p(\xc|x_s)\) and are averaged, to produce the summarized effect:

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation at this
level does not shed light to possible heterogeneous effects behind the
averaged explanation. Therefore, we model the uncertainty of the local
effects at \(\mathcal{H}(x_s)\) as the variance of the local
explanations:

@@

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
\end{equation}

and

>>>>

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), whereas to simple variables with plain lowercase
\( x \). Bold denotes a vector; \( \xb \) for simple variables or
\(\mathbf{X}\) for rvs. Often, we partition the input vector
\(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
feature effect of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the feature effect method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} There are three well-known feature
effect methods: PDPlots, MPlots and ALE. PDPlots formulate the feature effect of the \(s\)-th
attribute as an expectation over the marginal distribution
\(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\). MPlots formulate it
as an expectation over the conditional \(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as an
accumulation (integration) of the expected value of the local
effects:
\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
corelated input features. In this cases, PDPlots integrate over
unrealistic instances, due to the use of the marginal distribution
\( p(\mathbf{X}_c) \), and MPlots compute aggregated effects, i.e.,
impute the combined effect of sets of features to a single
feature. ALE manages to resolve both issues, and is therefore the only
trustable method in cases of correlated features.

\paragraph{Quantify the Heterogeneous Effects.}

Feature effect methods answer the question \textit{what happens
  (effect) to the output, if I increase/decrease the value of a
  specific feature}. Having answered the question above, it comes
naturally to also wonder \textit{how certain we are about the change
  on the output}. For this reason, a lot of interest is given lately
for quantifying the level of uncertainty, along with the expected
effect. The level of uncertainty is mostly quantified by measuring the
existence of heterogeneous effects, i.e. whether there are local
explanations that deviate from the expected global effect. ICE and
d-ICE plots provide a visual understanding of the heterogeneous
effects on top of PDPs. Another approach targets on grouping the
heterogeneous effects, e.g., allocating ICE plots in homogeneous
clusters, by dividing the input space. Some other approaches, like
H-Statistic, Greenwel, move a step behind and try to quantify the
level of interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index indicates the possibility of
heterogeneous effects. The aforementioned approaches face two
pathogenies; They either do not quantify the uncertainty of the
feature effect directly or they are based on PDPs, and, therefore,
they are subject to the failure modes of PDPs in cases of correlated
features. To the best of our knowledge, no method so far targets on
quantify the uncertainty of the feature effect as it is modelled by
ALE.

\paragraph{Cluster Instances with homogeneous effects.}

In real ML scenarios, the expected feature effect and the uncertainty
are estimated from the limited instances of the training set. ALE
approximation requires an additional step, where the axis of the
\(s\)-th feature is split into a sequence of non-overlaping bins and a
single effect (expectation and uncertainty) is computed for from the
population of instances that lie inside each
bin. \citep{apley2020visualizing} proposed estimating the local
effects in each bin by evaluating the black box-funtion at the bin
limits:

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}

In contrast, (cite) proposed the Differential ALE (DALE) estimation
for quantifying the local effects on the training-set instances,
instead of the bin limits:

\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, it allows the recomputation
of the accumulated effect with different bin-splitting with near-zero
computational overhead. However, none of the approximations above
deals with the crucial problem of the optimal bin-spliting. They split
the axis in a blind-way, partition in \(K\) equally-sized which can
lead to erroneous approximations.

Instead, we propose treating the axis-spliting step as an unsupervised
clustering problem. The objective of the clustering problem should
fulfill in the best way to contradictory ojectives. First, secure
robust estimations of the expected effect and the uncertainty inside
each bin given the limited instances of the training set and, second,
create bins with as homogeneous local effects as possible, for not
losing fine-grain resolution feature effects due to wide bins.

@@

ALE defines the local effect of the \(s\)-th feature on \(f(\cdot)\)
at point \((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local
explanations at \(x_s\) are, then, weigthed by the conditional
distribution \(p(\xc|x_s)\) and are averaged, to produce the
summarized effect at \(x_s\):

\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
\noindent
As described at the Introduction, limiting the explanation to the
exepected value level does not shed light to possible heterogeneous
effects behind the averaged explanation. Therefore, we model the
uncertainty of the local effects at \(\mathcal{H}(x_s)\) as the
variance of the local explanations:

@@

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}_{\mu}(x_s) = \Delta x \sum_{k=1}^{k_x} \hat{\mu}(z_{k-1}, z_k)
\end{equation}
and
